---
title: "583_proposal"
author: "Avishek Saha and Noman Mohammad"
date: "2023-02-24"
output: 
  html_document:
    css: style.css
---

The data for this project was uploaded to [kaggle](https://www.kaggle.com/datasets/ravi72munde/uber-lyft-cab-prices) by the user RaviMunde and licensed under CC0: Public Domain, which allows for unrestricted use and distribution of the data. The data set we are analyzing includes simulated ride prices for Uber and Lyft, which were obtained through custom application queries collected using their respective API's. In parallel, additional data was also collected for weather conditions corresponding to the source destination for the cab rides. The data was collected for the interval from 2018-11-25 to 2018-12-18. A custom application was built in Scala to query data at regular intervals of 5 minutes for cab ride estimates and 1 hour for weather data. The data was then saved to DynamoDB. The chosen interval was to query as much data as possible without unnecessary redundancy. Finally, the data collected is stored into two separate CSV files: cab_rides.csv and weather.csv

### Statistical Description

```{r, echo=FALSE}

cab <- read.csv('cab_rides.csv')
weather <- read.csv('weather.csv')

``` 

The cab_rides.csv consists of 10 variables and 693,071 rows:

    |- cab_type : String
    |- destination : String
    |- distance: Float
    |- id: String
    |- name: String
    |- price: Float
    |- product_id: String
    |- source: String
    |- surge_multiplier: String
    |- time_stamp:Long

- `cab_type`: This variable represents the type of ride hailing service used, such as Uber or Lyft.
- `destination`: This variable represents the drop off location of the ride.
- `distance`: This variable represents the distance traveled from the pickup location to the drop off location in miles ranging from 0.02 to 7.86.
- `id`: This variable represents a unique identifier for each ride.
- `name`: This variable represents the type of ride service used, such as Lyft XL or UberX.
- `price`: This variable represents the cost of the ride in US dollars ranging from 2.5 to 97.5.
- `product_id`: This variable represents a unique identifier for the type of ride service used.
- `source`: This variable represents the pickup location of the ride.
- `surge_multiplier`: This variable represents how much the ride price was multiplied based on current ride demand ranging from 1 to 3.
- `time_stamp`: This variable represents the time of the ride as an epoch time stamp in seconds.

```{r}

# Summary of the cab dataset
summary(cab)

```

The weather.csv consists of 8 variables and 6276 rows:

    |- clouds : Float
    |- humidity : Float
    |- time_stamp : Float
    |- location : String
    |- temp : String
    |- pressure : Float
    |- wind : Float
    |- rain : Float
    
- `clouds`: This variable represents the percentage of cloud cover in the sky at the time the weather was recorded ranging from 0 to 1.
- `humidity`: This variable represents the percentage of humidity in the air at the time the weather was recorded ranging from 0.45 to 0.99.
- `time_stamp`: This variable represents the time the weather was recorded as an epoch time stamp in seconds.
- `location`: This variable represents the location where the weather was recorded. This corresponds to the source location in the cab data.
- `temp`: This variable represents the temperature in Fahrenheit at the time the weather was recorded ranging from 19.62 to 55.41.
- `pressure`: This variable represents the air pressure in millibars at the time the weather was recorded ranging from 988.40 to 1035.12.
- `wind`: This variable represents the wind speed in miles per hour at the time the weather was recorded ranging from 0.30 to 18.18.
- `rain`: This variable represents the amount of rain in inches for the last hour at the time the weather was recorded ranging from 0.0002 to 0.2740.

```{r}

# Summary of the weather dataset
summary(weather)

``` 

Additionally, it may be worth noting that the data collected may have some limitations and underlying scientific processes. For instance, the data may not fully represent the patterns of ride-sharing app usage and demand across all Boston locations due to the data collected not being publicly available and was simulated, it may only represent a glimpse of the demand and supply of rides during the collected time interval. Moreover, The demand for ride-sharing services is heavily influenced by human behavior, such as commuting patterns, weekend vs weekday travel, and response to weather conditions. These factors may be difficult to predict or model accurately, which could impact the quality of the data collected.

### Hypothesis

We aim to answer several scientific questions related to services such as Uber and Lyft using a mix of exploratory data analysis and developing predictive models learnt throughout the MDS program. We will investigate the price variation between Uber and Lyft, assuming they have the same starting and ending points and travel at the same time. We will also attempt to predict the fare of a ride based on the ride-hailing service provider, time of day, and weather conditions. Additionally, we will investigate the surge pricing multiplier for Uber and Lyft rides in Boston based on variables such as time of day, day of the week, and weather conditions. Finally, we will also explore whether the source and destination of the ride have a significant impact on cab prices when controlling for distance.Our analysis could help us understand the relationship between weather conditions and cab services, providing insights into how these services could be optimized during different weather conditions.